Dynamics in Algorithmic Recourse

### 2022-04-26 dev/submissions/aies/extended_abstract/extended_abstract.qmd
more work on AIES

---
title: Dynamics in Algorithmic Recourse
subtitle: Trustworthy Artificial Intelligence for Finance and Economics
format:
    pdf:
        documentclass: acmconf
        number-sections: true
        include-in-header: preamble.tex
        keep-tex: true
---

@@

Recent advances in artificial intelligence (AI) have propelled its adoption in domains outside of computer science including health care, bioinformatics and genetics. In finance, economics and other social sciences, applications of AI are still relatively limited. Decision-making in these fields has traditionally been guided by Generalized Linear Models (GLM), which are theoretically founded, interpretable and often sufficient to model relationships between variables. Model interpretability is crucial in the social sciences context, because inference is typically at least as important as predictive performance. Decision-makers in the social sciences are also typically required to explain their decisions to human stakeholders: central bankers, for example, are held accountable by the public for the policies they decide on. It is therefore not surprising that practitioners and academics in these fields are reluctant to adopt AI technologies that ultimately cannot be trusted. Deep learning models, for example, are generally considered as black boxes and therefore difficult to apply in a context that demands explanations. This PhD project is focused on exploring and developing methodologies that improve the trustworthiness of AI and thereby enable its application in Finance and Economics.

@@

**Counterfactual explanations** (CE) explain how inputs into a model need to change for it to produce different outputs. They are intuitive, simple and intrinsically linked to the potential outcome framework for causal inference, which social scientists are familiar with. Counterfactual explanations that involve realistic and actionable changes can be used for the purpose of **Algorithmic Recourse** (AR) to help individuals facing adverse decisions. An example relevant to the Finance and Econonomics domain is consumer credit: in this context AR can be used to guide individuals to improve their credit worthiness, should they have previously been denied access to credit based on a black-box decision-making system .

Existing work on CE and AR has largely been limited to the static setting: given some classifier $M: \mathcal{X} \mapsto \mathcal{Y}$ we are interested in finding close [@wachter2017counterfactual], actionable [@ustun2019actionable], plausible [@joshi2019towards, @antoran2020getting, @schut2021generating], sparse [@schut2021generating], diverse [@mothilal2020explaining] and ideally causally founded counterfactual explanations [@karimi2021algorithmic] for some individual $x$. The ability of counterfactual explanations to handle dynamics like data and model shifts remains a largely unexplored research challenge at this point [@verma2020counterfactual]. Only one recent work considers the implications of **exogenous** domain shifts on the validity of recourse [@upadhyay2021towards]. The authors propose a simple minimax objective, that minimizes the counterfactual loss function for a maximal domain and model shift. They show that their approach yields more robust counterfactuals than existing approaches.

This project investigates **endogenous** domain and model shifts, that is shifts that occur when AR is actually impemented by a proportion of individuals and the classifier is updated in response. @fig-dynamics illustrates this idea for a binary problem involving a probabilistic classifier and a greedy counterfactual generator proposed by @schut2021generating: AR leads to a domain shift, which in turn causes a drastic model shift. As this game of implementing AR and updating the classifier is repeated, the decision boundary moves in the opposite direction of the original training samples in the target class. We consider several aspects of these dynamics as problematic. Firstly, as the decision boundary moves in the direction of the non-target class, counterfactual paths become shorter: in the loan example, individuals that previously would have been denied credit based on their input features are suddenly considered as credit worthy. Average default risk across all borrowers can therefore be expected to increase. Conversely, lenders that anticipate or detect these sorts of dynamics may choose to still deny credit to individuals that have implemented AR, thereby comprimising the validity of AR.

To the best of my knowledge this is the first work investigating endogenous dynamic in algorithmic recourse. In future experiments I want to investigate how this phenomenon plays out across different benchmark datasets (including German credit [@germancredit1994], Boston Housing [@boston1993] and COMPAS [@compas2016]). Furthermore, I want to asses to what extent the magnitude and direction of domain and model shifts depends on the choice of the counterfactual generator. To this end, I am currently supervising a group of undergraduate students, who are tackling some of these tasks in their final-year research project.

![PLACEHOLDER: The dynamics of algorithmic recourse.](www/dynamics.png){#fig-dynamics fig.pos="h" width=250px}

@@

To ensure that the generated explanations are realistic it is important to understand which input-output pairs are likely and which are not. To quantify their joint likelihood, previous work has either relied on generative models or restricted the analysis to probabilistic models that incorporate uncertainty in their predictions. While the former approach is more versatile since it is applicable to both deterministic and probabilistic models, the latter is computationally much more efficient. The approach proposed by @schut2021generating and used to generate the examples in @fig-dynamics falls into the latter category. In future work I want to explore how recent advances in post-hoc uncertainty quantification, most notably Laplace Redux [@daxberger2021laplace], can be leveraged to generate realistic and unambiguous counterfactual explanations for any model.^[For some initial work on this see my Julia implementation of Laplace Redux: [BayesLaplace.jl](https://www.paltmeyer.com/BayesLaplace.jl/dev/).]

### XAI for Time Series Models

Datasets in Finance and Economics often involve autoregressive time series, which is modelled differently than static, cross-sectional data. One popular approach to macroeconomic forecasting, for example, is Vector Autoregression (VAR), which involves an interpretable system of linear equations. Recent work has shown that machine learning can be leveraged in this context to improve forecasting performance, albeit at the sacrifice of explainability [@altmeyer2021deep].^[For an open-source implementation of our approach see [deepvars](https://github.com/pat-alt/deepvars): an R package Deep Vector Autoregression]. Through my PhD research I hope to ultimately adress that shortfall through explainable AI and probabilistic methods.

#### Counterfactual Explanations

I would like to investigate if and how CE can be used to leverage machine learning for time series forecasting without sacrificing explainability. In the context of conventional VAR counterfactual outcomes are actually well-studied [@kilian2017structural], which makes macroeconomic forecasting and interesting case study. More generally though, work on applying CE to black-box models for time series data has so far been limited and therefore warrants attention.

#### Probabilistic Methods

I am also interested in seeing if and how Laplace Redux can be applied to reccurent neural networks.

>>>>

---
title: Dynamics in Algorithmic Recourse
subtitle: Trustworthy Artificial Intelligence for Finance and Economics
format:
    pdf:
        documentclass: acmart
        classoption: sigconf
        number-sections: true
        include-in-header: preamble.tex
        keep-tex: true
---

@@

Recent advances in artificial intelligence (AI) have propelled its adoption in domains outside of computer science including health care, bioinformatics and genetics. In finance, economics and other social sciences, applications of AI are still relatively limited. Decision-making in these fields has traditionally been guided by interpretable models that facilitate explanations. Explainability is crucial in the social sciences context, because inference is typically at least as important as predictive performance. Decision-makers in the social sciences are also typically required to explain their decisions to human stakeholders: central bankers, for example, are held accountable by the public for the policies they decide on. It is therefore not surprising that practitioners and academics in these fields are reluctant to adopt AI technologies that cannot be explained. Deep neural networks, for example, are generally considered as black boxes and therefore not trustworthy in a context that demands explanations. This PhD project is focused on exploring and developing methodologies that improve the trustworthiness of AI and thereby enable its application in Finance and Economics.

@@

**Counterfactual explanations** (CE) explain how inputs into a model need to change for it to produce different outputs. They are intuitive, simple and intrinsically linked to the potential outcome framework for causal inference, which social scientists are familiar with. Counterfactual explanations that involve realistic and actionable changes can be used for the purpose of **Algorithmic Recourse** (AR) to help individuals facing adverse decisions. An example relevant to the Finance and Economics domain is consumer credit: in this context AR can be used to guide individuals to improve their credit worthiness, should they have previously been denied access to credit based on a black-box decision-making system .

Existing work on CE and AR has largely been limited to the static setting: given some classifier $M: \mathcal{X} \mapsto \mathcal{Y}$ we are interested in finding close [@wachter2017counterfactual], actionable [@ustun2019actionable], realistic [@joshi2019towards, @schut2021generating], sparse, diverse [@mothilal2020explaining] and ideally causally founded counterfactual explanations [@karimi2021algorithmic] for some individual $x$. The ability of counterfactual explanations to handle dynamics like data and model shifts remains a largely unexplored research challenge at this point [@verma2020counterfactual]. Only one recent work considers the implications of **exogenous** domain shifts on the validity of recourse [@upadhyay2021towards]. The authors propose a simple minimax objective, that minimizes the counterfactual loss function for a maximal domain and model shift. They show that their approach yields more robust counterfactuals than existing approaches.

This project investigates **endogenous** domain and model shifts, that is shifts that occur when AR is actually implemented by a proportion of individuals and the classifier is updated in response. @fig-dynamics illustrates this idea for a binary problem involving a probabilistic classifier and a greedy counterfactual generator proposed by @schut2021generating: AR leads to a domain shift, which in turn causes a drastic model shift. As this game of implementing AR and updating the classifier is repeated, the decision boundary moves in the opposite direction of the original training samples in the target class. We consider several aspects of these dynamics as problematic. Firstly, as the decision boundary moves in the direction of the non-target class, counterfactual paths become shorter: in the loan example, individuals that previously would have been denied credit based on their input features are suddenly considered as credit worthy. Average default risk across all borrowers can therefore be expected to increase. Conversely, lenders that anticipate or detect these sorts of dynamics may choose to still deny credit to individuals that have implemented AR, thereby comprimising the validity of AR.

To the best of my knowledge this is the first work investigating endogenous dynamic in algorithmic recourse. In future experiments I want to investigate how this phenomenon plays out across different benchmark datasets including German credit, Boston Housing and COMPAS. Furthermore, I want to assess to what extent the magnitude and direction of domain and model shifts depends on the choice of the counterfactual generator. To this end, I am currently supervising a group of undergraduate students, who are tackling some of these tasks in their final-year research project.

![PLACEHOLDER: The dynamics of algorithmic recourse.](){#fig-dynamics fig.pos="h" height=100px}

@@

To ensure that the generated explanations are realistic it is important to understand which input-output pairs are likely and which are not. With respect to the work-in-progress presented here, I expect that this may help in mitigating endogenous domain and model shifts. To quantify the joint likelihood of input-output pairs, previous work has either relied on generative models or restricted the analysis to probabilistic models that incorporate uncertainty in their predictions. While the former approach is more versatile, since it is applicable to both deterministic and probabilistic models, the latter is computationally much more efficient. The approach proposed by @schut2021generating and used to generate the examples in @fig-dynamics falls into the latter category.  In future work, I want to explore how recent advances in post-hoc uncertainty quantification, most notably Laplace Redux [@daxberger2021laplace], can be leveraged to generate realistic and unambiguous counterfactual explanations for any model.^[For some initial work on this see my Julia implementation of Laplace Redux: [BayesLaplace.jl](https://www.paltmeyer.com/BayesLaplace.jl/dev/).]

### 2022-04-26 dev/submissions/aies/extended_abstract/extended_abstract.tex
more work on AIES

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{acmconf}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Dynamics in Algorithmic Recourse},
  pdfauthor={Patrick Altmeyer},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\ConferenceShortName{AIES '22}
\ConferenceName{Fifth AAAI/ACM Conference on Artificial Intelligence, Ethics, and Society}
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\title{Dynamics in Algorithmic Recourse}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{Trustworthy Artificial Intelligence for Finance and Economics}
\author{Patrick Altmeyer}
\date{April 25, 2022}

@@

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, breakable, sharp corners, interior hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, frame hidden]}{\end{tcolorbox}}\fi

@@

\textbf{Counterfactual explanations} (CE) explain how inputs into a
model need to change for it to produce different outputs. They are
intuitive, simple and intrinsically linked to the potential outcome
framework for causal inference, which social scientists are familiar
with. Counterfactual explanations that involve realistic and actionable
changes can be used for the purpose of \textbf{Algorithmic Recourse}
(AR) to help individuals facing adverse decisions. An example relevant
to the Finance and Econonomics domain is consumer credit: in this
context AR can be used to guide individuals to improve their credit
worthiness, should they have previously been denied access to credit
based on a black-box decision-making system .

Existing work on CE and AR has largely been limited to the static
setting: given some classifier \(M: \mathcal{X} \mapsto \mathcal{Y}\) we
are interested in finding close (Wachter, Mittelstadt, and Russell
2017), actionable (Ustun, Spangher, and Liu 2019), plausible Schut et
al. (2021), sparse (Schut et al. 2021), diverse (Mothilal, Sharma, and
Tan 2020) and ideally causally founded counterfactual explanations
(Karimi, Schölkopf, and Valera 2021) for some individual \(x\). The
ability of counterfactual explanations to handle dynamics like data and
model shifts remains a largely unexplored research challenge at this
point (Verma, Dickerson, and Hines 2020). Only one recent work considers
the implications of \textbf{exogenous} domain shifts on the validity of
recourse (Upadhyay, Joshi, and Lakkaraju 2021). The authors propose a
simple minimax objective, that minimizes the counterfactual loss
function for a maximal domain and model shift. They show that their
approach yields more robust counterfactuals than existing approaches.

This project investigates \textbf{endogenous} domain and model shifts,
that is shifts that occur when AR is actually impemented by a proportion
of individuals and the classifier is updated in response.
Figure~\ref{fig-dynamics} illustrates this idea for a binary problem
involving a probabilistic classifier and a greedy counterfactual
generator proposed by Schut et al. (2021): AR leads to a domain shift,
which in turn causes a drastic model shift. As this game of implementing
AR and updating the classifier is repeated, the decision boundary moves
in the opposite direction of the original training samples in the target
class. We consider several aspects of these dynamics as problematic.
Firstly, as the decision boundary moves in the direction of the
non-target class, counterfactual paths become shorter: in the loan
example, individuals that previously would have been denied credit based
on their input features are suddenly considered as credit worthy.
Average default risk across all borrowers can therefore be expected to
increase. Conversely, lenders that anticipate or detect these sorts of
dynamics may choose to still deny credit to individuals that have
implemented AR, thereby comprimising the validity of AR.

To the best of my knowledge this is the first work investigating
endogenous dynamic in algorithmic recourse. In future experiments I want
to investigate how this phenomenon plays out across different benchmark
datasets (including German credit ({``German Credit Data''} 1994),
Boston Housing ({``Boston Housing Data''} 1993) and COMPAS (Larson et
al. 2016)). Furthermore, I want to asses to what extent the magnitude
and direction of domain and model shifts depends on the choice of the
counterfactual generator. To this end, I am currently supervising a
group of undergraduate students, who are tackling some of these tasks in
their final-year research project.

@@

{\centering \includegraphics[width=2.60417in,height=\textheight]{www/dynamics.png}

@@

To ensure that the generated explanations are realistic it is important
to understand which input-output pairs are likely and which are not. To
quantify their joint likelihood, previous work has either relied on
generative models or restricted the analysis to probabilistic models
that incorporate uncertainty in their predictions. While the former
approach is more versatile since it is applicable to both deterministic
and probabilistic models, the latter is computationally much more
efficient. The approach proposed by Schut et al. (2021) and used to
generate the examples in Figure~\ref{fig-dynamics} falls into the latter
category. In future work I want to explore how recent advances in
post-hoc uncertainty quantification, most notably Laplace Redux
(Daxberger et al. 2021), can be leveraged to generate realistic and
unambiguous counterfactual explanations for any model.\footnote{For some
  initial work on this see my Julia implementation of Laplace Redux:
  \href{https://www.paltmeyer.com/BayesLaplace.jl/dev/}{BayesLaplace.jl}.}

\hypertarget{xai-for-time-series-models}{%
\subsection{XAI for time series
models}\label{xai-for-time-series-models}}

Datasets in Finance and Economics often involve autoregressive time
series, which is modelled differently than static, cross-sectional data.
One popular approach to macroeconomic forecasting, for example, is
Vector Autoregression (VAR), which involves an interpretable system of
linear equations. Recent work has shown that machine learning can be
leveraged in this context to improve forecasting performance, albeit at
the sacrifice of explainability (Altmeyer, Agusti, and Vidal-Quadras
Costa 2021).\footnote{For an open-source implementation of our approach
  see \href{https://github.com/pat-alt/deepvars}{deepvars}: an R package
  Deep Vector Autoregression}. Through my PhD research I hope to
ultimately adress that shortfall through explainable AI and
probabilistic methods.

\hypertarget{counterfactual-explanations}{%
\subsubsection{Counterfactual
Explanations}\label{counterfactual-explanations}}

I would like to investigate if and how CE can be used to leverage
machine learning for time series forecasting without sacrificing
explainability. In the context of conventional VAR counterfactual
outcomes are actually well-studied (Kilian and Lütkepohl 2017), which
makes macroeconomic forecasting and interesting case study. More
generally though, work on applying CE to black-box models for time
series data has so far been limited and therefore warrants attention.

\hypertarget{probabilistic-methods}{%
\subsubsection{Probabilistic Methods}\label{probabilistic-methods}}

I am also interested in seeing if and how Laplace approximation can be
used for Bayesian deep learning with time series data. I hope that the
findings from both of these projects can ultimately be used to build
complex but interpretable time series models for classification and
forecasting in finance and economics.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-altmeyer2021deep}{}}%
Altmeyer, Patrick, Marc Agusti, and Ignacio Vidal-Quadras Costa. 2021.
{``Deep Vector Autoregression for Macroeconomic Data.''}
\url{https://thevoice.bse.eu/wp-content/uploads/2021/07/ds21-project-agusti-et-al.pdf}.

\leavevmode\vadjust pre{\hypertarget{ref-antoran2020getting}{}}%
Antorán, Javier, Umang Bhatt, Tameem Adel, Adrian Weller, and José
Miguel Hernández-Lobato. 2020. {``Getting a Clue: A Method for
Explaining Uncertainty Estimates.''} \emph{arXiv Preprint
arXiv:2006.06848}.

\leavevmode\vadjust pre{\hypertarget{ref-boston1993}{}}%
{``Boston Housing Data.''} 1993.
\url{https://archive.ics.uci.edu/ml/machine-learning-databases/housing/}.

\leavevmode\vadjust pre{\hypertarget{ref-daxberger2021laplace}{}}%
Daxberger, Erik, Agustinus Kristiadi, Alexander Immer, Runa Eschenhagen,
Matthias Bauer, and Philipp Hennig. 2021. {``Laplace Redux-Effortless
Bayesian Deep Learning.''} \emph{Advances in Neural Information
Processing Systems} 34.

\leavevmode\vadjust pre{\hypertarget{ref-germancredit1994}{}}%
{``German Credit Data.''} 1994.
\url{https://archive.ics.uci.edu/ml/datasets/statlog+(german+credit+data)}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-kilian2017structural}{}}%
Kilian, Lutz, and Helmut Lütkepohl. 2017. \emph{Structural Vector
Autoregressive Analysis}. Cambridge University Press.

\leavevmode\vadjust pre{\hypertarget{ref-compas2016}{}}%
Larson, Jeff, Surya Mattu, Lauren Kirchner, and Julia Angwin. 2016.
{``COMPAS.''} \url{https://github.com/propublica/compas-analysis}.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  sigconf]{acmart}
\usepackage{amsmath}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Dynamics in Algorithmic Recourse},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

% \ConferenceShortName{AIES '22}
% \ConferenceName{Fifth AAAI/ACM Conference on Artificial Intelligence, Ethics, and Society}

\setcopyright{acmcopyright}
\copyrightyear{2022}
\acmYear{2022}
\acmDOI{XXXXXXX.XXXXXXX}

\acmConference[AIES '22]{Fifth AAAI/ACM Conference on Artificial Intelligence, Ethics, and Society}{Oxford, UK}

\author{Patrick Altmeyer}
\affiliation{%
  \institution{Delft University of Technology}
  \city{Delft}
  \country{The Netherlands}
}
\email{p.altmeyer@tudelft.nl}
\orcid{0000-0003-4726-8613}
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\title{Dynamics in Algorithmic Recourse}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{Trustworthy Artificial Intelligence for Finance and Economics}
\author{}
\date{April 26, 2022}

@@

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, frame hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, sharp corners, interior hidden, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

\textbf{Counterfactual explanations} (CE) explain how inputs into a
model need to change for it to produce different outputs. They are
intuitive, simple and intrinsically linked to the potential outcome
framework for causal inference, which social scientists are familiar
with. Counterfactual explanations that involve realistic and actionable
changes can be used for the purpose of \textbf{Algorithmic Recourse}
(AR) to help individuals facing adverse decisions. An example relevant
to the Finance and Economics domain is consumer credit: in this context
AR can be used to guide individuals to improve their credit worthiness,
should they have previously been denied access to credit based on a
black-box decision-making system .

Existing work on CE and AR has largely been limited to the static
setting: given some classifier \(M: \mathcal{X} \mapsto \mathcal{Y}\) we
are interested in finding close (Wachter, Mittelstadt, and Russell
2017), actionable (Ustun, Spangher, and Liu 2019), realistic Schut et
al. (2021), sparse, diverse (Mothilal, Sharma, and Tan 2020) and ideally
causally founded counterfactual explanations (Karimi, Schölkopf, and
Valera 2021) for some individual \(x\). The ability of counterfactual
explanations to handle dynamics like data and model shifts remains a
largely unexplored research challenge at this point (Verma, Dickerson,
and Hines 2020). Only one recent work considers the implications of
\textbf{exogenous} domain shifts on the validity of recourse (Upadhyay,
Joshi, and Lakkaraju 2021). The authors propose a simple minimax
objective, that minimizes the counterfactual loss function for a maximal
domain and model shift. They show that their approach yields more robust
counterfactuals than existing approaches.

This project investigates \textbf{endogenous} domain and model shifts,
that is shifts that occur when AR is actually implemented by a
proportion of individuals and the classifier is updated in response.
Figure~\ref{fig-dynamics} illustrates this idea for a binary problem
involving a probabilistic classifier and a greedy counterfactual
generator proposed by Schut et al. (2021): AR leads to a domain shift,
which in turn causes a drastic model shift. As this game of implementing
AR and updating the classifier is repeated, the decision boundary moves
in the opposite direction of the original training samples in the target
class. We consider several aspects of these dynamics as problematic.
Firstly, as the decision boundary moves in the direction of the
non-target class, counterfactual paths become shorter: in the loan
example, individuals that previously would have been denied credit based
on their input features are suddenly considered as credit worthy.
Average default risk across all borrowers can therefore be expected to
increase. Conversely, lenders that anticipate or detect these sorts of
dynamics may choose to still deny credit to individuals that have
implemented AR, thereby comprimising the validity of AR.

To the best of my knowledge this is the first work investigating
endogenous dynamic in algorithmic recourse. In future experiments I want
to investigate how this phenomenon plays out across different benchmark
datasets including German credit, Boston Housing and COMPAS.
Furthermore, I want to assess to what extent the magnitude and direction
of domain and model shifts depends on the choice of the counterfactual
generator. To this end, I am currently supervising a group of
undergraduate students, who are tackling some of these tasks in their
final-year research project.

@@

{\centering \includegraphics[width=\textwidth,height=1.04167in]{.pdf}

@@

To ensure that the generated explanations are realistic it is important
to understand which input-output pairs are likely and which are not.
With respect to the work-in-progress presented here, I expect that this
may help in mitigating endogenous domain and model shifts. To quantify
the joint likelihood of input-output pairs, previous work has either
relied on generative models or restricted the analysis to probabilistic
models that incorporate uncertainty in their predictions. While the
former approach is more versatile, since it is applicable to both
deterministic and probabilistic models, the latter is computationally
much more efficient. The approach proposed by Schut et al. (2021) and
used to generate the examples in Figure~\ref{fig-dynamics} falls into
the latter category. In future work, I want to explore how recent
advances in post-hoc uncertainty quantification, most notably Laplace
Redux (Daxberger et al. 2021), can be leveraged to generate realistic
and unambiguous counterfactual explanations for any model.\footnote{For
  some initial work on this see my Julia implementation of Laplace
  Redux:
  \href{https://www.paltmeyer.com/BayesLaplace.jl/dev/}{BayesLaplace.jl}.}

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-daxberger2021laplace}{}}%
Daxberger, Erik, Agustinus Kristiadi, Alexander Immer, Runa Eschenhagen,
Matthias Bauer, and Philipp Hennig. 2021. {``Laplace Redux-Effortless
Bayesian Deep Learning.''} \emph{Advances in Neural Information
Processing Systems} 34.

@@
